---
title: "Class 05: Data Visualization with ggplot2"
author: Pagna Hout
date: 2023-04-19
format: pdf
editor_options: 
  chunk_output_type: inline
---

Load package

```{r}
library(ggplot2)
ggplot(cars)
```

Specifying aesthetic mappings with aes ()

```{r}
ggplot(cars) +
  aes (x=speed, y=dist)
```

Specifying a geom layer with geom_point()

```{r}
ggplot(cars) +
  aes (x=speed, y=dist) +
  geom_point()
```

### Adding trend line layer to help show the relationship between the plot variables with the geom_smooth() function

```{r}
ggplot(cars) +
  aes (x=speed, y=dist) +
  geom_point() +
  geom_smooth()
```

### Adding straight line from linear model without the shaded standard error region

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE)
```

### Adding various label annotations with the labs() function and changing the plot to look more conservative "black & white" theme by adding the theme.bw() function

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) +
  geom_point() +
  labs(title="Speed and Stopping Distances of Cars",
       x="Speed (MPH)", 
       y="Stopping Distance (ft)",
       subtitle = "Your informative subtitle text here",
       caption="Dataset: 'cars'") +
  geom_smooth(method="lm", se=FALSE) +
  theme_bw()
```

### Adding more plot aesthetics through aes()

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

```{r}
nrow(genes)
ncol(genes)
colnames(genes)

table(genes$State)
round (table(genes$State)/nrow(genes)*100,2)
```

First basic scatter plot of the dataset:

```{r}
ggplot(genes) + 
    aes(x=Condition1, y=Condition2) +
    geom_point()
```

There is extra information in this dataset, namely the `State` column, which tells us whether the difference in expression values between conditions is statistically significant. Let's map this column to point color:

```{r}
p <- ggplot(genes) + 
    aes(x=Condition1, y=Condition2, col=State) +
    geom_point()
p
```

```{r}
p + scale_colour_manual( values=c("blue","gray","red") )
```

```{r}
p + scale_colour_manual(values=c("blue","gray","red")) +
    labs(title="Gene Expresion Changes Upon Drug Treatment",
         x="Control (no drug) ",
         y="Drug Treatment")
```
