---
title: "Class 06: R functions"
author: Pagna Hout
date: 2023-04-21
format: pdf
editor_options: 
  chunk_output_type: inline
---

In this class, we will develop our own R function to calculate average grades in a fictional class.

We will start with a simplified version of the problem, just calculating the average grade of one student.

```{r}
# Example input vectors to start with
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)

```

We are going to start by calculating the average score of the homework.

```{r}
mean(student1)
```

To get the minimum score, we can use which.min.

```{r}
student1
which.min(student1)
```

I can do the average of the first 7 homework scores.

```{r}
mean(student1[1:7])
```

Another way to select the first 7 homework scores:

```{r}
student1[1:7]
student1[-8]
```

Another way to drop the lowest score (and storing a vector in another variable to use later):

```{r}
student1_drop_lowest = student1[-which.min(student1)]
student1_drop_lowest
```

To get the mean with the lowest score dropped:

```{r}
mean(student1_drop_lowest)
```

We have our first working snippet of code!

Let's try to generalize it to student 2

```{r}
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
student2_drop_lowest = student2[-which.min(student2)]
student2_drop_lowest
```

There is a way to calculate the mean dropping missing values (or NA).

```{r}
mean(student2, na.rm = TRUE)
```

This looks good for student2. However, for student3...

```{r}
mean(student3, na.rm=TRUE)
```

We want to know the position of the NAs. So, for student2, we can use the following:

```{r}
which(is.na(student2))
```

For student3:

```{r}
which(is.na(student3))
```

For considering missing values, we can mask the NA with zeros.

Student2:

```{r}
student2[is.na(student2)] <- 0
student2
```

If I use the same code for student3:

```{r}
student3[is.na(student3)] <- 0
student3
```

This is going to be our final working snippet of code for all students (with and without NA values)

```{r}
student3[is.na(student3)] <- 0
student3_drop_lowest = student3[-which.min(student3)]
mean(student3_drop_lowest)
```

Let's build a function:

```{r}
function(x){
  x[is.na(x)] <- 0
  x_drop_lowest = x[-which.min(x)]
  mean(x_drop_lowest)
}
```

We can now write it as a function:

```{r}
#' Calculate the average score for a vector of 
#' homework scores, dropping the lowest score, 
#' and considering NA values as zeros.
#'
#' @param x A numeric vector of scores
#'
#' @return The average value of homework scores
#' @export
#'
#' @examples
#' 
#' student <- c('100', '50', NA)
#' grade(student)
#' 
grade <- function(x){
  # Mask NA values with 0
  x[is.na(x)] <- 0
  # Dropping lowest score
  x_drop_lowest = x[-which.min(x)]
  mean(x_drop_lowest)
}
```

Let's apply the function to the students' grades

```{r}
grade(student1)
grade(student2)
grade(student3)
```

Let's apply our function to a gradebook from this URL: \"https://tinyurl.com/gradeinput\"

```{r}
URL <- "https://tinyurl.com/gradeinput"
gradebook <- read.csv(URL, row.names=1)
```

Let's apply my function grade to the gradebook using apply and running it by rows using MARGIN=1

```{r}
apply(gradebook, MARGIN=1, grade)
```

Q2. Using your grade() function and the supplied gradebook, who is the top scoring student overall in the gradebook?

```{r}
max(apply(gradebook, 1, grade))
```

The maximum score is 94.5.

```{r}
which.max(apply(gradebook, 1, grade))
```

The student getting the maximum overall score was student 18.

Q3. From your analysis of the gradebook, which homework was toughest on students (i.e. obtained the lowest scores overall?)

```{r}
#masking NA values with 0
gradebook [is.na(gradebook)] <- 0
```

Now we apply the mean function to the gradebook

```{r}
apply(gradebook, 2, mean)
```

The toughest homework will be homework 2 considering the mean and considering missing homework as 0.

Maybe having zeros for missing homework is too strict and is not a good representation of the homework difficulty.

One thing we can do is to remove the missing values.

```{r}
gradebook <- read.csv(URL, row.names=1)
apply(gradebook, 2, mean, na.rm = TRUE)
```

Instead of assignment zeros to missing values, if we directly don't consider missing values, the toughest homework will be homework 3 (according to the mean).

If we use the median instead of the mean as a measure of overall score...

```{r}
apply(gradebook, 2, median, na.rm = TRUE)
```

Using some plots...

```{r}
boxplot(gradebook)
```

Q4. From your analysis of the gradebook, which homework was most predictive of overall score (i.e. highest correlation with average grade score)?

```{r}
overall_grades = apply (gradebook, 1, grade)
overall_grades
```

```{r}
cor(gradebook$hw1, overall_grades)
```

```{r}
gradebook [is.na(gradebook)] <- 0
apply(gradebook, 2, cor, y = overall_grades)
```

The maximum value is ...

```{r}
which.max (apply(gradebook, 2, cor, y = overall_grades))
```

Homework 5 was the most predictive of overall score.
