---
title: "Class 10: Halloween Mini-Project"
author: Pagna Hout
date: 2023-05-05
format: pdf
---

## Importing candy data

```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```

#### Q1: How many different candy types are in this dataset?

```{r}
dim(candy)
```

There are 85 different candy types in this data set.

#### Q2: How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

There are 38 fruity candy types in this data set.

## What is your favorite candy?

#### Q3: What is your favorite candy in the dataset and what is it's `winpercent` value?

```{r}
candy["Haribo Happy Cola", ]$winpercent
```

My favorite candy is Haribo Happy Cola and its winpercent value is 34.15896.

#### Q4: What is the `winpercent` value for "Kit Kat"?

```{r}
candy["Kit Kat",]$winpercent
```

The winpercent value for "Kit Kat" is 76.7686.

#### Q5: What is the `winpercent` value for "Tootsie Roll Snack Bars"?

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```

The winpercent value for "Tootsie Roll Snack Bars" is 49.6535.

```{r}
library("skimr")
skim(candy)
```

#### Q6: Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The winpercent variable seems to be on a different scale.

#### Q7: What do you think a zero and one represent for the `candy$chocolate` column?

I think a zero represents "no chocolate" and a one means that there is chocolate in the candy.

#### Q8: Plot a histogram of `winpercent` values

```{r}
library(ggplot2)
hist(candy$winpercent)
```

#### Q9: Is the distribution of `winpercent` values symmetrical?

No, the distribution is not symmetrical.

#### Q10: Is the center of the distribution above or below 50%?

Yes

```{r}
t.test(candy$winpercent[as.logical(candy$chocolate)], candy$winpercent[as.logical(candy$fruity)])
```

#### Q11: On average is chocolate candy higher or lower ranked than fruit candy?

On average, chocolate candy is higher ranked than fruity candy.

#### Q12: Is this difference statistically significant?

Yes, this different is statistically significant.

## Overall Candy Rankings

```{r}
library(dplyr)
candy %>% 
  arrange(winpercent) %>% 
  head(5)

candy %>%
  arrange(winpercent) %>%
  tail(5)
  
```

#### Q13: What are the five least liked candy types in this set?

The five least liked candies are Nik K Nip, Boston Baked Beans, Chiclets, Super Bubble and Jawbusters.

#### Q14: What are the top 5 all time favorite candy types out of this set?

The top 5 all time favorites are Reese's Peanut Butter Cups, Reese's Miniatures, Twix, Kit Kat and Snickers.

#### Q15: Make a first barplot of candy ranking based on `winpercent` values.

```{r}
library(ggplot2)

ggplot(candy,aes(winpercent, rownames(candy))) +
  geom_col()
```

#### Q16: This is quite ugly, use the `reorder()` function to get the bars sorted by `winpercent`?

```{r}
ggplot(candy,aes(winpercent, reorder(rownames(candy),winpercent))) +
  geom_col()
```

Adding some colors...

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

#### Q17: What is the worst ranked chocolate candy?

Sixlets

#### Q18: What is the best ranked fruity candy?

Starburst

## Taking a look at pricepercent

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 70)
```

#### Q19: Which candy type is the highest ranked in terms of `winpercent` for the least money - i.e. offers the most bang for your buck?

Reese's Peanut Butter Cup

#### Q20: What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

Nik L Nip, Nestle Smarties, Ring Pop, Hershey's Krackel, and Hershey's Milk Chocolate

## Exploring the correlation structure

```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```

#### Q22: Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and fruity

#### Q23: Similarly, what two variables are most positively correlated?

Bar and pluribus

## Principle Component Analysis

```{r}
pca <- prcomp(candy, scale=T)
summary(pca)
```

```{r}
plot(pca$x[,1:2])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 70)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
library(plotly)

ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

#### Q24: What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, hard and pluribus.
